Reducing Acceptance Marks in Emerson-Lei Automata by QBF Solving

### 2023-06-04 paper/main.tex
Updated delag data in paper.
moving tables and figures in Section 6
Added heatmap2
polishing mainly in Section 6

\begin{abstract}
  This paper presents a novel application of QBF solving to automata
  reduction. We focus on \emph{Transition-based Emerson-Lei automata
    (TELA)}, which is a popular formalism that generalizes many
  traditional kinds of automata over infinite words including Büchi,
  co-Büchi, Rabin, Streett, and parity automata. Transitions in a TELA
  are labelled with acceptance marks and its accepting formula is a
  positive Boolean combination of atoms saying that a particular mark
  has to be visited infinitely or finitely
  often. % in an accepting run.
  Algorithms processing these automata are often very sensitive to the
  number of %different
  acceptance marks. We introduce a new technique for reducing the
  number of %different
  acceptance marks in TELA based on satisfiability of \emph{quantified
    Boolean formulas (QBF)}.
  % : if a constructed QBF formula is satisfiable then
  % its model encodes an equivalent accepting formula with less
  % acceptance marks.
  %
  % In fact, we present three formula constructions
  % % trading speed for higher reduction efficiency.
  % with increasing reduction potential and formula complexity, and thus
  % also decreasing speed of their satisfiability solving.
  We evaluated
  our reduction technique on TELA produced by state-of-the-art tools
  of the libraries Owl and Spot and by the tool \texttt{ltl3tela}. The
  technique reduced some acceptance marks in automata produced by all
  the tools. \fixme{On automata with more than one acceptance mark produced
  by tools Delag and Rabinizer~4 of the Owl library, our technique reduced
  29.1\% and 38.5\% of acceptance marks, respectively.}
  % The achieved
  % reduction of acceptance marks on automata produced by tools of the Owl library was
  % between 25 and 40\%.
\end{abstract}

@@

The presented reduction algorithms have been implemented in a tool
called \telatko. It is implemented in Python 3 and uses the Spot
library~\cite{duret.16.atva2} for automata parsing and manipulation,
and the theorem prover Z3~\cite{demoura.08.tacas} to solve the
satisfiability of QBF transformed to prenex normal form (and not using
CNF). Our tool is available at
\begin{center}
\url{https://gitlab.fi.muni.cz/xschwar3/telatko}
\end{center}
under the GNU GPLv3
license. The tool can be executed by the command 
\begin{center}
  \texttt{telatko -F <input.hoa> [-L }$j$\texttt{] [-C] [-I] [-T }$t$\texttt{] [-O <output.hoa>]}
\end{center}
where
\begin{description}
\item[\texttt{-F <input.hoa>}] specifies the file with the input automaton in HOA format~\cite{babiak.15.cav}, 
\item[\texttt{-L }$j$] specifies the reduction level; if omitted, the multi-level reduction is used,
\item[\texttt{-C}] switches on the reduction of the number of cubes
  after the number of marks is reduced (it corresponds to
  $\mathit{reduceC}=\true$ in Algorithms~\ref{alg:single}
  and~\ref{alg:multi}),
\item[\texttt{-I}] switches on the incremental version,
\item[\texttt{-T }$t$] sets the timeout for each QBF query to $t$ seconds (the default value is 50 seconds), 
\item[\texttt{-O <output.hoa>}] specifies the output file; if omitted,
  the produced automaton is sent to \textit{stdout} in the HOA format.
\end{description}

@@

To evaluate our reduction technique, we applied \telatko to automata
produced by the following process. We started with two sets of LTL formulas.
\begin{itemize}
\item One set contains all LTL formulas from \emph{literature} that
  are provided by the tool \texttt{genltl} of the Spot
  library~\cite{duret.16.atva2}~2.10.4. For parametrized formula
  patterns, we consider instances for all combinations of parameter
  values from 1 to 4.
\item The second set consists of 400 random LTL formulas with 4 atomic
  propositions. These formulas were generated by the tool
  \texttt{randltl} of the Spot library.
\end{itemize}
On both these sets, we applied the tool \texttt{ltlfilt} of the Spot
library to simplify the formulas and remove duplicates and formulas
equivalent to $\true$ and $\false$. After these steps, we had 348
LTL formulas from literature and 335 random formulas.
%
Formulas from both sets have been translated to nondeterministic TELA
by two state-of-the-art translators, namely \texttt{ltl2tgba} (used
with option \texttt{-G}) from the Spot library~\cite{duret.16.atva2}
and \texttt{ltl3tela}~\cite{major.19.atva}, and to deterministic TELA
by \texttt{ltl3tela} with option \texttt{-D1} and by another two
state-of-the-art translators from the Owl
library~\cite{kretinsky.18.atva}, namely \texttt{ltl2dela} (known as
Delag)~\cite{muller.17.gandalf} and \texttt{ltl2dgra} (known as
Rabinizer~4)~\cite{kretinsky.18.cav}. Some translators failed on some
formulas: it usually reached a timeout of 60 seconds or it produced an
automaton that cannot be parsed by the Spot library.
% produced by the individual tools contains less than 348 automata as not
% all of the formulas are successfully translated due to a timeout of 60
% seconds.
Further, we have removed automata with 0 or 1 acceptance mark as
there is a little point in reducing these. Table
\ref{tab:ltltranslators} shows the exact versions of the
translators. For each set of formulas and for each translator, it also
provides the number of fails, the number of produced automata with
less than two marks, and the number of automata with at least two
marks. The numbers of automata with at least two marks are typeset in
bold as these automata are actually used for the experimental
evaluation of our reduction technique.

@@

For each automata set identified by the translator and the set of
formulas, Table~\ref{tab:cumm-data} shows the cumulative numbers of
marks in the input automata sets and after each reduction, together
with the reduction ratio and total time spent by the considered
reduction. The column \textit{solver timeout} shows the number of
automata for which the last query to QBF solver did not finish within
the 30~seconds limit. The timeout of the last QBF query means that the
automaton may be potentially further reduced if a longer time limit is
used. \fixme{Zkontrolovat zbytek odstavce podle aktualnich vysledku,
  specifikovat formula set.} One can observe that a higher level
sometimes achieves a smaller reduction than a lower level (e.g.,
compare level~1 and level~2 for \texttt{ltl2dgra}). This is caused by
the QBF solver timeouts occuring earlier as formulas constructed by
the higher level are more complex. For \texttt{ltl2dela} and
\texttt{ltl2dgra}, the automata sets do not contain any automaton
where level~2 or level~3 achieves a better result than
level~1. However, all levels contribute to the reductions in the
multi-level setting.

\begin{table}[t]

@@

@@

  & 332 & 4.6~~~ & 465.9 & 12~~~
  & 334 & 4.0~~~ & 553.9 & 15~~~
  & 326 & 6.3~~~ & 691.8 & 18~~~
  & 323 & 7.2~~~ & 1545.9 & 18~~~

% \\
% level 1 &
% 387~ & 26.0~~~~~~ & 755.4~ &  18~~~
% \\
% &
% level 2 &
% 390~ & 25.4~~~~~~ & 1024.5~ &  20~~~
% \\
% &
% level 3 &
% 393~ & 24.9~~~~~~ & 7273.3~ &  25~~~
% \\
% &
% multi &
% 371~ & 29.1~~~~~~ & 9111.4~ &  21~~~
% \\
% \midrule
% \multirow{4}{*}{\makecell{\texttt{ltl2dgra}\\234 automata\\882 acc.~marks}}
% &
% level 1 &
% 552~ & 37.4~~~~~~ & 738.7~ & 14~~~
% \\
% &
% level 2 &
% 569~ & 35.5~~~~~~ & 907.3~ &  18~~~
% \\
% &
% level 3 &
% 561~ & 36.4~~~~~~ & 1163.4~ &  21~~~
% \\
% &
% multi &
% 542~ & 38.5~~~~~~ & 2320.7~ &  21~~~
% \\
% \midrule
% \multirow{4}{*}{\makecell{\texttt{ltl3tela}\\91 automata\\348 acc.~marks}}
% &
% level 1 &
% 332~ & 4.6~~~~~~ & 465.9~ &  12~~~
% \\
% &
% level 2 &
% 334~ & 4.0~~~~~~ & 553.9~ &  15~~~
% \\
% &
% level 3 &
% 326~ & 6.3~~~~~~ & 691.8~ &  18~~~
% \\
% &
% multi &
% 323~ & 7.2~~~~~~ & 1545.9~ & 18~~~
% \\
\bottomrule                                                                     
\end{tabular}
\end{table}

\fixme{Upravit zbytek textu.}
Table~\ref{tab:heatmap} shows the number of automata from all
four sets that have a given number of original acceptance marks and a
given number of acceptance marks after multi-level reduction. The
table indicates that in many cases only 1 or 2~marks can be
saved. However, the achieved reduction is substantial for some
automata with a higher number of original acceptance marks. For
example, in 24 cases, we have reduced 7 or more
acceptance marks to only 4.
% Moreover, there were many timeouts of the last QBF query for manof these automata.
% Let us note that some that some reductions are the results are promising, e.g. in 17 cases, we have successfully
% reduced 8 or more acceptance marks to only 4.

\begin{table}[b!]
\caption{The effect of multi-level reduction.
A cell on coordinates $(x,y)$ contains the number of automata that have been
reduced from $x$ to $y$ acceptance marks. If the cell contains a sum of two
numbers, the latter represents the number of automata where the attempt to reduce
another mark has been unsuccessful due to a QBF solver timeout.}
\label{tab:heatmap}
\setlength{\tabcolsep}{2.5pt}
\centering
\input{table-heatmap.tex}
\end{table}

>>>>

\begin{abstract}
  This paper presents a novel application of QBF solving to automata
  reduction. We focus on \emph{Transition-based Emerson-Lei automata
    (TELA)}, which is a popular formalism that generalizes many
  traditional kinds of automata over infinite words including Büchi,
  co-Büchi, Rabin, Streett, and parity automata. Transitions in a TELA
  are labelled with acceptance marks and its accepting formula is a
  positive Boolean combination of atoms saying that a particular mark
  has to be visited infinitely or finitely
  often. % in an accepting run.
  Algorithms processing these automata are often very sensitive to the
  number of %different
  acceptance marks. We introduce a new technique for reducing the
  number of %different
  acceptance marks in TELA based on satisfiability of \emph{quantified
    Boolean formulas (QBF)}.
  % : if a constructed QBF formula is satisfiable then
  % its model encodes an equivalent accepting formula with less
  % acceptance marks.
  %
  % In fact, we present three formula constructions
  % % trading speed for higher reduction efficiency.
  % with increasing reduction potential and formula complexity, and thus
  % also decreasing speed of their satisfiability solving.
  We evaluated our reduction technique on TELA produced by
  state-of-the-art tools of the libraries Owl and Spot and by the tool
  \texttt{ltl3tela}. The technique reduced some acceptance marks in
  automata produced by all the tools. On automata with more than one
  acceptance mark obtained by translation of LTL formulas from
  literature with tools Delag and Rabinizer~4, our
  technique reduced 27.7\% and 39.3\% of acceptance marks,
  respectively. The reduction was even higher on automata from random
  formulas.
  % The achieved
  % reduction of acceptance marks on automata produced by tools of the Owl library was
  % between 25 and 40\%.
\end{abstract}

@@

The presented reduction algorithms have been implemented in a tool
called \telatko. It is implemented in Python 3 and uses the Spot
library~\cite{duret.16.atva2} for automata parsing and manipulation,
and the theorem prover Z3~\cite{demoura.08.tacas} to solve the
satisfiability of QBF transformed to prenex (non-CNF) normal form. Our
tool is available at
\begin{center}
\url{https://gitlab.fi.muni.cz/xschwar3/telatko}
\end{center}
under the GNU GPLv3
license. The tool can be executed by the command 
\begin{center}
  \texttt{telatko -F <input.hoa> [-L }$j$\texttt{] [-C] [-I] [-T }$t$\texttt{] [-O <output.hoa>]}
\end{center}
where
\begin{description}
\item[\texttt{-F <input.hoa>}] specifies the file with the input automaton in HOA format~\cite{babiak.15.cav}, 
\item[\texttt{-L }$j$] specifies the reduction level; if omitted, the multi-level reduction is used,
\item[\texttt{-C}] switches on the reduction of the number of cubes
  after the number of marks is reduced (it corresponds to
  $\mathit{reduceC}=\true$ in Algorithms~\ref{alg:single}
  and~\ref{alg:multi}),
\item[\texttt{-I}] switches on the incremental version,
\item[\texttt{-T }$t$] sets the timeout for each QBF query to $t$ seconds (the default value is 50 seconds), 
\item[\texttt{-O <output.hoa>}] specifies the output file; if omitted,
  the produced automaton is sent to \textit{stdout} in the HOA format.
\end{description}

@@

To evaluate our reduction technique, we applied \telatko to automata
produced by the following process. We started with two sets of LTL formulas.
\begin{itemize}
\item One set contains all LTL formulas from \emph{literature} that
  are provided by the tool \texttt{genltl} of the Spot
  library~\cite{duret.16.atva2}~2.10.4. For parametrized formula
  patterns, we consider instances for all combinations of parameter
  values from 1 to 4.
\item The second set consists of 400 random LTL formulas with 4 atomic
  propositions. These formulas were generated by the tool
  \texttt{randltl} of the Spot library.
\end{itemize}
On both these sets, we applied the tool \texttt{ltlfilt} of the Spot
library to simplify the formulas and remove duplicates and formulas
equivalent to $\true$ and $\false$. After these steps, we had 348
LTL formulas from literature and 335 random formulas.
%
Formulas from both sets have been translated to nondeterministic TELA
by two state-of-the-art translators, namely \texttt{ltl2tgba} (used
with option \texttt{-G}) from the Spot library~\cite{duret.16.atva2}
and \texttt{ltl3tela}~\cite{major.19.atva}, and to deterministic TELA
by \texttt{ltl3tela} with option \texttt{-D1} and by two
state-of-the-art translators from the Owl
library~\cite{kretinsky.18.atva}, namely \texttt{ltl2dela} (known as
Delag)~\cite{muller.17.gandalf} and \texttt{ltl2dgra} (known as
Rabinizer~4)~\cite{kretinsky.18.cav}. Some translators failed on some
formulas: it usually reached a timeout of 60 seconds or it produced an
automaton that cannot be parsed by the Spot library.
% produced by the individual tools contains less than 348 automata as not
% all of the formulas are successfully translated due to a timeout of 60
% seconds.
Further, we have removed automata with 0 or 1 acceptance mark as there
is a little point in reducing these. Table \ref{tab:ltltranslators}
shows the exact versions of the translators. For each translator and
each set of formulas, the table also provides the number of fails, the
number of produced automata with less than two marks, and the number
of automata with at least two marks. The numbers of automata with at
least two marks are typeset in bold as these automata are actually
used for the experimental evaluation of our reduction technique.

@@

\begin{table}[tp]

@@

  & \multicolumn{4}{c}{\mybox{523 marks in 129 automata}}
  & 386 & 26.2~~~ & 811.4 & 18~~~
  & 391 & 25.2~~~ & 1071.6 & 19~~~
  & 397 & 24.1~~~ & 7326.8 & 26~~~
  & 378 & 27.7~~~ & 9186.0 & 24~~~

@@

  \bottomrule                                                                     
\end{tabular}
\end{table}

For each automata set identified by the translator and the set of
formulas, Table~\ref{tab:cumm-data} shows the cumulative numbers of
marks in the input automata set and after each reduction, together
with the reduction ratio and total time spent by the considered
reduction. The column \textit{solver timeout} shows the number of
automata for which the last query to QBF solver did not finish within
the 30~seconds limit. The timeout of the last QBF query means that the
automaton may be potentially further reduced if a longer time limit is
used. One can observe that a higher level sometimes achieves a smaller
reduction than a lower level (e.g., compare level~1 and level~2 for
\texttt{ltl3tela} on automata coming from formulas from
literature). This is caused by the QBF solver timeouts occuring
earlier as formulas constructed by the higher level are more
complex. The automata sets produced by \texttt{ltl2dela} and
\texttt{ltl2dgra}, on formulas from literature do not contain any
automaton where level~2 or level~3 achieves a better result than
level~1. However, all levels contribute to the reductions in the
multi-level setting.

\fixme{Upravit zbytek textu a popis Table~3.}
Table~\ref{tab:heatmap} shows the number of automata from all
four sets that have a given number of original acceptance marks and a
given number of acceptance marks after multi-level reduction. The
table indicates that in many cases only 1 or 2~marks can be
saved. However, the achieved reduction is substantial for some
automata with a higher number of original acceptance marks. For
example, in 24 cases, we have reduced 7 or more
acceptance marks to only 4.
% Moreover, there were many timeouts of the last QBF query for manof these automata.
% Let us note that some that some reductions are the results are promising, e.g. in 17 cases, we have successfully
% reduced 8 or more acceptance marks to only 4.

\begin{table}[t]
\caption{The effect of multi-level reduction.
A cell on coordinates $(x,y)$ contains the number of automata that have been
reduced from $x$ to $y$ acceptance marks. If the cell contains a sum of two
numbers, the latter represents the number of automata where the attempt to reduce
another mark has been unsuccessful due to a QBF solver timeout.}
\label{tab:heatmap}
\setlength{\tabcolsep}{2.5pt}
\centering
\input{table-heatmap2.tex}
\end{table}
